# Mini Project 01: Fiscal Characteristics of Major US Public Transit System

## Introduction

Public transit systems plays an important role in providing the essential transportation service for millions of residents in the United States. In this project, we are looking to analyze the fiscal characteristics of public transit systems to evaluate their financial performance. 

## Fiscal Characteristics:
We are going to analyze public transit systems based on the following:

1. **Expenses:** costs to operate (fuel, maintenance, employee salaries, etc)

2. **Fare revenue:** money earned from passenger fares

3. **Farebox recovery ratio:** how much of the operating expense is covered by the fares revenue (total farewell revenue/total operating expense)

4. **UPT:** Unlinked Passenger Trips

5. **VRM:** Vehicle Revenue Miles

## Data source:

We used data from the [National Transit Database](https://www.transit.dot.gov/ntd). Data files required:

- [2022 Fare Revenue](http://www.transit.dot.gov/sites/fta.dot.gov/files/2024-04/2022%20Fare%20Revenue.xlsx)
- [2022 Operating Expenses](https://data.transportation.gov/api/views/dkxx-zjd6/rows.csv?date=20231102&accessType=DOWNLOAD&bom=true&format=true)
- [Monthly ridership](https://www.transit.dot.gov/sites/fta.dot.gov/files/2024-09/July%202024%20Complete%20Monthly%20Ridership%20%28with%20adjustments%20and%20estimates%29_240903.xlsx)

I created the table using the code provided in the assignment instructions. In this code, I renamed "UZA Name" to "metro_area" (Task 1). I also modified the "Mode" column (Task 2) so we can understand the context of each abbrevations. The interpretation of the codes can be found on the NTD website in the [glossary](https://www.transit.dot.gov/ntd/national-transit-database-ntd-glossary) section. A sample of 1000 observations is shown below:

```{r echo=FALSE, message=FALSE}
if(!require("tidyverse")) install.packages("tidyverse")
library(tidyverse)
fare_revenue_file <- "C://Users//zhuol//OneDrive//Documents//STA9750-2024-FALL//2022_fare_revenue.xlsx"

FARES <- readxl::read_xlsx(fare_revenue_file) |>
    select(-`State/Parent NTD ID`, 
           -`Reporter Type`,
           -`Reporting Module`,
           -`TOS`,
           -`Passenger Paid Fares`,
           -`Organization Paid Fares`) |>
    filter(`Expense Type` == "Funds Earned During Period") |>
    select(-`Expense Type`) |>
    group_by(`NTD ID`,       # Sum over different `TOS` for the same `Mode`
             `Agency Name`,  # These are direct operated and sub-contracted 
             `Mode`) |>      # of the same transit modality
                             # Not a big effect in most munis (significant DO
                             # tends to get rid of sub-contractors), but we'll sum
                             # to unify different passenger experiences
    summarize(`Total Fares` = sum(`Total Fares`)) |>
    ungroup()

expenses_file <- "C://Users//zhuol//OneDrive//Documents//STA9750-2024-FALL//2022_expenses.csv"

EXPENSES <- readr::read_csv(expenses_file) |>
  select(`NTD ID`, 
         `Agency`,
         `Total`, 
         `Mode`) |>
  mutate(`NTD ID` = as.integer(`NTD ID`)) |>
  rename(Expenses = Total) |>
  group_by(`NTD ID`, `Mode`) |>
  summarize(Expenses = sum(Expenses)) |>
  ungroup()

FINANCIALS <- inner_join(FARES, EXPENSES, join_by(`NTD ID`, `Mode`)) |>
   mutate(Mode=case_when( 
    Mode == "HR" ~ "Heavy Rail", 
    Mode == "DR" ~ "Demand Response",
    Mode == "FB" ~ "Ferry Boat",
    Mode == "MB" ~ "Bus",
    Mode == "SR" ~ "Streetcar Rail",
    Mode == "TB" ~ "Trolleybus",
    Mode == "VP" ~ "Vanpool",
    Mode == "CB" ~ "Commuter Bus",
    Mode == "RB" ~ "Bus Rapid Transit",
    Mode == "LR" ~ "Light Rail",
    Mode == "YR" ~ "Hybrid Rail",
    Mode == "MG" ~ "Monorail and Automated Guideway ",
    Mode == "CR" ~ "Commuter Rail",
    Mode == "AR" ~ "Alaska Railroad",
    Mode == "TR" ~ "Aerial Tramway",
    Mode == "IP" ~ "Inclined Plane",
    Mode == "PB" ~ "Publico",
    Mode == "CC" ~ "Cable Car",
    TRUE ~ "Unknown"))  #modified the mode column to make everything consistent with Task 2

# Monthly Transit Numbers
ridership_file <- "C://Users//zhuol//OneDrive//Documents//STA9750-2024-FALL//ridership.xlsx"

TRIPS <- readxl::read_xlsx(ridership_file, sheet="UPT") |>
  filter(`Mode/Type of Service Status` == "Active") |>
  select(-`Legacy NTD ID`, 
         -`Reporter Type`, 
         -`Mode/Type of Service Status`, 
         -`UACE CD`, 
         -`TOS`) |>
  pivot_longer(-c(`NTD ID`:`3 Mode`), 
               names_to="month", 
               values_to="UPT") |>
  drop_na() |>
  mutate(month=my(month)) # Parse _m_onth _y_ear date specs

MILES <- readxl::read_xlsx(ridership_file, sheet="VRM") |>
  filter(`Mode/Type of Service Status` == "Active") |>
  select(-`Legacy NTD ID`, 
         -`Reporter Type`, 
         -`Mode/Type of Service Status`, 
         -`UACE CD`, 
         -`TOS`) |>
  pivot_longer(-c(`NTD ID`:`3 Mode`), 
               names_to="month", 
               values_to="VRM") |>
  drop_na() |>
  group_by(`NTD ID`, `Agency`, `UZA Name`, 
           `Mode`, `3 Mode`, month) |>
  summarize(VRM = sum(VRM)) |>
  ungroup() |>
  mutate(month=my(month)) # Parse _m_onth _y_ear date specs

##join Trips and miles using the NDI ID
USAGE <- inner_join(TRIPS, MILES) |>
  mutate(`NTD ID` = as.integer(`NTD ID`)) |>
  rename(metro_area = `UZA Name`)  ##Task 1

## Task 2: Recoding Mode column

# Step 1: Find unique Mode codes
unique_modes <- USAGE |> 
    distinct(Mode)

# Step 2: Rename codes 
USAGE <- USAGE |>
  mutate(Mode=case_when( 
    Mode == "HR" ~ "Heavy Rail", 
    Mode == "DR" ~ "Demand Response",
    Mode == "FB" ~ "Ferry Boat",
    Mode == "MB" ~ "Bus",
    Mode == "SR" ~ "Streetcar Rail",
    Mode == "TB" ~ "Trolleybus",
    Mode == "VP" ~ "Vanpool",
    Mode == "CB" ~ "Commuter Bus",
    Mode == "RB" ~ "Bus Rapid Transit",
    Mode == "LR" ~ "Light Rail",
    Mode == "YR" ~ "Hybrid Rail",
    Mode == "MG" ~ "Monorail and Automated Guideway ",
    Mode == "CR" ~ "Commuter Rail",
    Mode == "AR" ~ "Alaska Railroad",
    Mode == "TR" ~ "Aerial Tramway",
    Mode == "IP" ~ "Inclined Plane",
    Mode == "PB" ~ "Publico",
    Mode == "CC" ~ "Cable Car",
    TRUE ~ "Unknown"))  

if(!require("DT")) install.packages("DT")
library(DT)
library(gt)

sample_n(USAGE, 1000) |> 
  select(-`NTD ID`, -`3 Mode`) |>
  rename(Unlinked_Passenge_Ttrips = UPT, Vehicle_Revenue_Miles = VRM) |>
  mutate(month=as.character(month)) |>
  DT::datatable()  #visualize a sample of 1000 observations
```

Note: In this sample, I unselected columns "NTD ID" and "3 Mode" and renamed "UPT" and "VRM" to make the table look cleaner and easier to interpret.

## Task 3: Using `dplyr` to Analyze Transit Agencies

**1. What transit agency had the most total VRM in this sample?**

The MTA New York transit has the most vehicle mileage. It has operated over 10.8 billion (almost 11 billion) mileage.

``` {r}
most_vrm_agency <- USAGE |>
  group_by(Agency) |>
  summarize(total_vrm = sum(VRM, na.rm = TRUE)) |>
  ungroup() |>
  arrange(desc(total_vrm)) |>
  slice(1) 

most_vrm_agency |>
  gt() |>
  tab_header(title = "Most Total VRM Agency")
```

**2. What transit mode had the most total VRM in this sample?**

Bus has the most total VRM of over 49 billion mileage. 

```{r}
most_vrm_mode <- USAGE |>
  group_by(Mode) |>
  summarize(total_vrm = sum(VRM, na.rm = TRUE)) |>
  ungroup() |>
  arrange(desc(total_vrm)) |>
  slice(1)
most_vrm_mode |>
  gt() |>
  tab_header(title = "Most Total VRM Mode")
```

**3. How many trips were taken on the NYC Subway (Heavy Rail) in May 2024?**

In May 2024, a total of 180,458,819 trips were taken on the NYC Subway (Heavy Rail). 

```{r}
nyc_subway_trips_may_2024 <- USAGE |>
  mutate(month = format(ymd(month), "%Y-%m")) |>
  filter(Mode == "Heavy Rail",
         Agency == "MTA New York City Transit", 
         month == "2024-05") |>
  summarize(total_trips = sum(UPT, na.rm = TRUE)) 
nyc_subway_trips_may_2024 |>
  gt() |>
  tab_header(title = "NYC Subway (Heavy Rail) in May 2024")

```

**5. How much did NYC subway ridership fall between April 2019 and April 2020?**

Between April 2019 and April 2020, NYC Subway ridership fall about 91.28%. 

```{r}
nyc_subway_ridership_april_2019_to_april_2020 <- USAGE |>
  mutate(month = format(ymd(month), "%Y-%m")) |>
  filter(Agency == "MTA New York City Transit", Mode == "Heavy Rail", month %in% c("2019-04", "2020-04")) |>
   summarise(
    April_2019 = sum(case_when(month == "2019-04" ~ UPT, TRUE ~ 0), na.rm = TRUE),
    April_2020 = sum(case_when(month == "2020-04" ~ UPT, TRUE ~ 0), na.rm = TRUE), percent_change = (April_2020 - April_2019)/April_2019 *100) |>
  ungroup()

nyc_subway_ridership_april_2019_to_april_2020 |>
  gt() |>
  tab_header(title = "% Change in NYC Subway Ridership")
```

## Task 4: Explore and Analyze

### Monthly Ridership Trends

I wanted to explore the monthly ridership trends of the public transit systems. Using the code below, I made a table that lists the top 5 months with the greatest percent change in monthly ridership. 

```{r}
library(dplyr)
library(lubridate)
monthly_ridership_change <- USAGE |>
  mutate(month = format(ymd(month), "%Y-%m")) |>
  group_by(month) |>
  summarize(total_upt = sum(UPT, na.rm = TRUE), .groups = 'drop') |>
  arrange(month) |>
  mutate(percent_change = (total_upt - lag(total_upt)) / lag(total_upt) * 100)
# take the top 5 months that shows the greatest percent change

top_5_months <- monthly_ridership_change |>
  filter(!is.na(percent_change)) |>
  arrange(desc(percent_change)) |>
  slice_head(n = 5) 

top_5_months |>
  gt() |>
  tab_header(title = "Top 5 Months with Greatest Percent Change in Ridership")

```

**Findings:**

1. **2020-06 (29.76% Increase)** and **2020-07 (20.04% Increase)**: Due to quarantine requirements in early 2020, public transit usage has rapidly declined. In June 2020, as the restrictions loosened, and hence we see rebound in passenger usage of public transit with approximately 30% increase in usage. Passenger usage continue to grow a 20% increase in July 2020. Also, this growth can reflect seasonal factors since July is one of the peak seasons for summer vacation. People are regular to their regular routine and planning for vacations. 

2. **2021-03 (27.34% Increase)** and **2022-03 (20.37% Increase)**: These period reflects the ongoing recovery of the COVID-19 pandemic. Reopening of business, end of remote working modes, loosen of public health safety measures encourage more rider usage of public transit. Hence, we see a relatively big increase in ridership usage. 


### Unlinked Passenger Trips per Vehicle Revenue Miles

I also wanted to explore the ratio of unlinked passenger trips per vehicle revenue miles to evaluate the efficiency of transit systems. Using the code below, I categorized and ranked 18 transit modes by the UPT per VRM ratio. The UPT per VRM evaluates how many passenger trips is generated per vehicle mile. 

```{r}
UPT_per_VRM <- USAGE |>
  group_by(Mode) |>
  summarize(total_upt = sum(UPT, na.rm = TRUE), total_vrm = sum(VRM, na.rm = TRUE)) |>
  mutate(UPT_per_VRM = total_upt/total_vrm) |>
  arrange(desc(UPT_per_VRM))

UPT_per_VRM |>
  gt() |>
  tab_header(title = "Unlinked Passenger Trips per Vehicle Revenue Mile")

```

**Findings:**

1. **High Efficiency Modes:** 

        - Aerial Tramway
        - Inclined Planes
    
2. **Moderate Efficiency Modes:**

        - Ferry Boats
        - Cable Cars
    
3. **Lower Efficiency Modes: **

        - Demand Response
        - Bus
        - Commuter Bus

### Highest VRM Mode for Each Agency

Agencies have different transit modes. I want to analyze which transit mode has the highest UPT for each agency. For example, King County has transit modes of ferry boats, bus, demand response, etc. I want to explore which transit modes in King County generates the highest VRM. Using the code below, I generated the top 3 highest VRM transit mode for each agency. 

```{r}
top_3_highest_vrm_mode_per_agency <- USAGE |>
  group_by(Agency,Mode) |>
  summarize(total_vrm = sum(VRM, na.rm = TRUE), .groups = 'drop') |>
  arrange(desc(total_vrm)) |>
  slice_head(n=3)

top_3_highest_vrm_mode_per_agency |>
  gt() |>
  tab_header(title = "Highest VRM Mode for Each Agency") |>
  cols_label(total_vrm = "Total Vehicle Revenue Miles")
```
**Findings:**

The highest transit mode for MTA New York City Transit agency is heavy rail. It generates 7,732,916,753 miles. Recall in question 1 of Task 3, MTA New York City Transit has a total of 10,832,855,350 miles in total. If we calculate the percentage of total VRM operated by heavy rail, we get 71.4%. Heavy rails operates a big portion of the vehicle miles travelled by MTA.

## Task 5: 2022 Annual Summary Table

```{r}
##creating 2022 summary table
USAGE_2022_ANNUAL <- USAGE |>
  filter(year(month) == 2022) |>
  group_by(`NTD ID`, Agency, metro_area, Mode) |> 
  summarize(
    total_upt = sum(UPT, na.rm = TRUE),
    total_vrm = sum(VRM, na.rm = TRUE),
    .groups = 'drop') |>
  filter(total_upt > 400000) |> #filter major transit system
  ungroup()

```

Note: Since we are restricting our answers to major transit system, I have filtered total UPT to be greater than 400,000 (shown in code above). 

### Merging 2022 Summary Table with Financial Data Table

```{r}
##merge summary with Financials report
USAGE_AND_FINANCIALS <- left_join(USAGE_2022_ANNUAL, 
                                  FINANCIALS, 
                                  join_by(`NTD ID`, Mode)) |>
  drop_na()

```

## Task 6: Farebox Recovery Among Major Systems

**Note:** For Task 6 questions, the output table will show the top 5 results.

**1. Which transit system (agency and mode) had the most UPT in 2022?**

In 2022, the MTA New York City Transit heavy rail has the great number of passenger trips. It has over 1.7 billion trips. 

```{r}
most_total_upt <- USAGE_AND_FINANCIALS |>
  filter(total_upt >= 400000) |>
  arrange(desc(total_upt)) |>
  slice_head(n=5) |>
  select(Agency, Mode, total_upt)
most_total_upt |>
  gt() |>
  tab_header(title = "Transit System with the Most UPT in 2022")

```

**2. Which transit system (agency and mode) had the highest farebox recovery, defined as the highest ratio of Total Fares to Expenses?**

In 2022, Port Imperial Ferry Corporation ferry boat has the highest farebox recovery ratio of 1.428. 

```{r}
highest_farebox_recovery <- USAGE_AND_FINANCIALS |>
  mutate(farebox_recovery = `Total Fares`/Expenses) |>
  select(`NTD ID`, Agency, Mode, `Total Fares`, Expenses, farebox_recovery) |>
  arrange(desc(farebox_recovery)) |>
  slice_head(n=5) 
highest_farebox_recovery |>
  gt() |>
  tab_header(title = "Transit System with the Highest Farebox Recovery")
```

**3. Which transit system (agency and mode) has the lowest expenses per UPT?**

In 2022, North Carolina State University Bus has the lowest expense per UPT of $1.18. 

```{r}
lowest_expenses_per_upt <- USAGE_AND_FINANCIALS |>
  mutate(expenses_per_upt = Expenses/total_upt) |>
  arrange(expenses_per_upt) |>
  slice_head(n=5)
lowest_expenses_per_upt |>
  gt() |>
  tab_header(title = "Transit System with the Lowest Expenses per UPT")

```

**4. Which transit system (agency and mode) has the highest total fares per UPT?**

In 2022, commuter bus under agency Hampton Jitney Inc has the highest fares per UPT of $41.30.

```{r}
highest_fares_per_upt <- USAGE_AND_FINANCIALS |>
  mutate(fares_per_upt = `Total Fares`/total_upt) |>
  arrange (desc(fares_per_upt)) |>
  slice_head(n=5) 
highest_fares_per_upt |>
  gt() |>
  tab_header(title = "Transit System with the Highest Total Fares per UPT")

```

**5. Which transit system (agency and mode) has the lowest expenses per VRM?**

In 2022, Metropolitan Transportation Commission vanpool has the lowest expense per VRM of $0.45. 

```{r}
lowest_expenses_per_vrm <- USAGE_AND_FINANCIALS |>
  mutate(expenses_per_vrm = Expenses/total_vrm) |>
  select(-`NTD ID`,
         -metro_area, 
         -`Agency Name`) |>
  arrange(expenses_per_vrm) |>
  slice_head(n=5)
lowest_expenses_per_vrm |>
  gt() |>
  tab_header(title = "Transit System with the Lowest Expenses per VRM")

```

**6. Which transit system (agency and mode) has the highest total fares per VRM?**

In 2022, Jacksonville Transportation Authority ferryboat has the highest total fares per VRM of $157.70. 

```{r}
highest_fares_per_vrm <- USAGE_AND_FINANCIALS |>
  mutate(fares_per_vrm = `Total Fares`/total_vrm) |>
  select(-`NTD ID`,
         -metro_area,
         -`Agency Name`) |>
  arrange(desc(fares_per_vrm)) |>
  slice_head(n=5)
highest_fares_per_vrm |>
  gt() |>
  tab_header(title = "Transit System with the Highest Total Fares per VRM")
```

## Conclusion

The metrics to analyze the most efficient transit system can be measured in different ways. 

Looking at the **farebox recovery ratio**, Port Imperial Ferry Corporation ferry boat would be the top candidate as it can shows a strong ratio of 1.428 (refer back to Task 6 question 2). This shows that Port Imperial Ferry Corporation ferry boat has strong profitability and has a healthy financial performance. 

If we measure efficiency from **vehicle miles**,  MTA New York City Transit Heavy Rail would be the most efficient transit system. It has the highest vehicle miles travelled (refer to question 1 of Task 3). It provides extensive amount of service indicating high ridership demand. Its vehicles are well utilized and therefore operating efficiently.



